### 2019-06-04 2-Chapters/2-Chapter/chapter2.tex
Almost done for UCB proof, I'll then be VERY quick for klUCB

% - Fairness ?
Finally, another interesting measure of performance is the \emph{fairness}.
It was not studied much for single-player problems, and usually it means that each arm must be explored at least a given fraction of the total number of rounds $T$ (\ie, ``arm-wise fairness'').
It was studied in a very recent article \cite{Patil2019stochastic},
where a different notion of regret is introduced to account for the fairness constraint. The authors show that different algorithms, based on $\UCB_1$ or Thompson sampling, can achieve logarithmic regret while respecting the fairness constraints.
%
In this thesis, we only consider fairness in the multi-player bandit models,
in Chapter~\ref{chapter:5}, where fairness refers to a different notion (\ie, ``cooperative fairness'').
When $M<K$ players learn in a decentralized way, they will essentially converge to an orthogonal affectation of players to arms, so that each player uses a unique arm among the $M$ best arms (most of the times).
The fairness constraint says that, in average, no user is arbitrarily going to converge more frequently on a better arm than another user, and it is for instance the case for permutation invariant decentralized algorithms.
We refer to Section~\ref{sub:5:betterLowerBound} for more details.

@@

% ----------------------------------------------------------------------------
\subsection{Index policies, $\UCB_1$, \klUCB{} and others}

A large family of algorithms are index policies, that compute an index $U_k(t)$ on each arm $k$ at time $t$.
The indexes should represent the expected quality of each arm, thus index policies play the arm that maximizes their index, \ie, $A(t) = \argmax_k U_k(t)$.
If more than one indexes maximize $(U_k(t))_k$, the arm is chosen among the set, usually in a uniformly random manner: $A(t) \sim \cU(\argmax_k U_k(t))$.
%
The Algorithm~\ref{algo:2:indexPolicy} below details a generic index policy, that includes well known and efficient algorithms such as $\UCB_1$, \klUCB{} and many others.

@@

\textbf{The $\UCB_1$ index policy: using Hoeffding's inequality to build confidence intervals.}
%
Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t)=\frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) = \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}$.
%
\begin{align}\label{eq:UCB_index}
    U_k^{\UCB_1}(t) = \UCB_k(t) = \underbrace{\frac{X_k(t)}{N_k(t)}}_{\widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}}_{\xi_k(t)}.
\end{align}

This selection rule \eqref{eq:UCB_index} makes sense for the following reason:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons:
$i)$ because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ but $N_k(t)\ll\log(t)$, thus all arms must be explored at least $\mathcal{O}(\log(T))$ times.
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{}  respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems).

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>10$,
    the $\UCB_1$ algorithm verifies the following asymptotic regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \smallO{\log(T)}.
    \end{equation}
\end{theorem}
%
\begin{proof}
    For simplicity, we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality} as stated above in Lemma~\ref\label{lem:2:HoeffdingInequality}.
    To analyze the algorithm $\UCB_1$, we restrict to $\alpha=2$ and we proceed as follows:

    $i)$ First, for any arm $k$, Hoeffding's inequality shows that $\Pr(\UCB_k(t) \leq \mu_k) \leq \bigO{1/t^{4}}$ independently of $N_k(t)$,
    and so $\sum_{t=1}^T \Pr(\UCB_k(t) \leq \mu_k) = \smallO{\log(T)}$: the \UCB{} are almost always above the empirical means (\ie, they are true \emph{upper} confidence bounds).

    $ii)$
    Then, we denote again $k^*$ an optimal arm, and $A(t)$ is the arm chosen in round $t$.
    According to the algorithm, the probability that $k$ is chosen is upper-bounded by $\Pr(\UCB_k(t) \leq \UCB_{k^*}(t))$.
    This can happen if $\UCB_{k^*}(t) \leq \mu_{k^*}$, which from $i)$ is very unlikely, or if $\UCB_k(t) \leq \UCB_{k^*}(t) > \mu_{k^*}$.
    By writing $\mu_k = \mu_{k^*} - \Delta_k$, and because $\UCB_k(t)$ are using $N_k(t)$ \iid{} samples from arm $k$, of mean $\mu_k$, Hoeffding's inequality also shows that $\Pr(\UCB_k(t) - \mu_k > \Delta_k) \leq 8/(\Delta_k^2 t)$ (the confidence radius $\xi_k(t)$ is actually built exactly to obtain this bound).
    %
    And thus\footnote{~We used the partial sum of the harmonic series: $\sum\limits_{t=1}^T \frac{1}{t} \underset{T\to\infty}{\sim} \log(T)$.} $\sum_{t=1}^T \Pr(\UCB_k(t) - \mu_k > \Delta_k) \leq 1/\Delta_k^2 \sum_{t=1}^T 8/t = 1/\Delta_k^2 8\log(T)$.

    To sum up, we obtain the following bound on $N_k(T)$ for any sub-optimal arm $k$:
    \begin{align*}
        \E[N_k(T)] = \sum_{t=1}^T \Pr(\ind(A(t)=k)) &\leq \sum_{t=1}^T \Pr(\UCB_k(t) > \UCB_{k^*}(t)) \\
        &\leq \sum_{t=1}^T \Pr(\UCB_{k^*}(t) < \mu_{k^*}) + \sum_{t=1}^T \Pr(\UCB_{k^*}(t) < \mu_{k^*}) \\
        &\leq \smallO{\log(T)} + \frac{8}{\Delta_k^2} \log(T).
    \end{align*}
    %
    By using the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, this gives $R_T^{\UCB_1} \leq \sum_{k: \Delta_k>0} \frac{8}{\Delta_k} \log(T) + \smallO{\log(T)}$.
    We preferred to use the $\smallO{\dots}$ Landau notation, instead of keeping the exact constants, for simplicity and clarity.
    More details on the proofs can be found for instance in Chapter~2 of \cite{Bubeck12}, or Section~1.3 of \cite{Slivkins2019}.
\end{proof}

What this theorem shows is that the $\UCB_1$ algorithm achieves order-optimal problem-dependent bounds with a $\cO(K\log(T)/\Delta^2)$ regret.
The $\UCB_1$ algorithm has the advantage of not relying on any prior knowledge of the problem difficulty, of being anytime, and computationally simple, for both memory and time aspects.
The $\UCB_1$ algorithm requires a constant storage capacity for each arm (storing $X_k(t)$ and $N_k(t)$ require two integers or a \texttt{float} and an integer, \ie, $\bigO{1}$), giving a storage capacity of $\bigO{K}$, independently of the horizon $T$.
It also needs to perform some mathematical operations, but a square-root and a logarithm are cheap, thus it costs a $\bigO{1}$ constant time for each arm and time step, thus a time $\bigO{K T}$ for the $T$ rounds.
%
We use it as the base building block for other policies in Chapter~\ref{chapter:4}, for these reasons.
Moreover, the $\UCB_1$ algorithm has the advantage of being easy to explain and understand, and its decisions can be interpreted clearly: by reading the values of $\widehat{\mu_k}(t)$ and $\UCB_k(t)$, anyone can see why an arm was chosen at anytime.
This is why it is used in our demonstration presented in Section~\ref{sec:4:gnuradio}.

>>>>

% - Fairness ?
Finally, another interesting measure of performance is the \emph{fairness}.
It was not studied much for single-player problems, and usually it means that each arm must be explored at least a given fraction of the total horizon $T$ (\ie, ``arm-wise fairness'').
It was studied in a very recent article \cite{Patil2019stochastic},
where a different notion of regret is introduced to account for the fairness constraint. The authors show that different algorithms, based on $\UCB_1$ or Thompson sampling, can achieve logarithmic regret while respecting the fairness constraints.
%
In this thesis, we only consider fairness in the multi-player bandit models,
in Chapter~\ref{chapter:5}, where fairness refers to a different notion (\ie, ``cooperative fairness'').
When $M<K$ players learn in a decentralized way, they will essentially converge to an orthogonal affectation of players to arms, so that each player uses a unique arm among the $M$ best arms (most of the times).
The fairness constraint says that, in average, no user is arbitrarily going to converge more frequently on a better arm than another user, and it is for instance the case for permutation invariant decentralized algorithms.
We refer to Section~\ref{sub:5:betterLowerBound} for more details.

@@

% ----------------------------------------------------------------------------
\subsection{Index policies: $\UCB_1$, \klUCB{} and others}

A large family of algorithms are index based, as they compute an index $U_k(t)$ on each arm $k$ at time $t$.
The indexes should represent the expected quality of each arm, thus index policies play the arm that maximizes their index, \ie, $A(t) = \argmax_k U_k(t)$.
If more than one indexes maximize $(U_k(t))_k$, the arm is chosen among the set, usually in a uniformly random manner: $A(t) \sim \cU(\argmax_k U_k(t))$.
%
The Algorithm~\ref{algo:2:indexPolicy} below details a generic index policy, that includes well known and efficient algorithms such as $\UCB_1$, \klUCB{} and many others.

@@

\textbf{The $\UCB_1$ index policy: using Hoeffding's inequality to build confidence intervals.}\\
%
Let us consider another approach for adaptive exploration, known as ``optimism under uncertainty'': assume each arm is as good as it can possibly be given the observations so far, and choose the best arm based on these optimistic estimate.
%
This intuition leads to the $\UCB_1$ algorithm, initially introduced in \cite{Auer02}.
For a parameter $\alpha$, the \UCB{} indexes are computed as follows, as a sum of
the \emph{average reward} $\widehat{\mu_k}(t)=\frac{X_k(t)}{N_k(t)}$
and a \emph{confidence radius} $\xi_k(t) = \sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}$.
%
\begin{align}\label{eq:UCB_index}
    U_k^{\UCB_1}(t) = \UCB_k(t) = \underbrace{\frac{X_k(t)}{N_k(t)}}_{\widehat{\mu_k}(t)} + \underbrace{\sqrt{\alpha \frac{\log\left(t\right)}{N_k(t)}\right)}}_{\xi_k(t)}.
\end{align}

This selection rule \eqref{eq:UCB_index} makes sense for the following reason:
an arm $k$ is chosen at step $t$ because it has a large $\UCB_k(t)$, which can happen for two reasons:
$i)$ because the average reward $\widehat{\mu_k}(t)$ is large, in which case this arm is likely to have a high mean reward $\mu_k$,
$ii)$ and/or because the confidence radius $\xi_k(t)$ is large, in which case this arm has not been explored much.
Either reason makes this arm worth choosing.
One can also observe that $\xi_k(t)\to\infty$ if $t\to\infty$ while $N_k(t)\ll\log(t)$, which can be used to prove that the $\UCB_1$ algorithm tries all arms at least $\mathcal{O}(\log(T))$ times (in average).
%
Furthermore, these two terms $\widehat{\mu_k}(t)$ and $\xi_k(t)$ in the \UCB{}  respectively represent \emph{exploitation} and \emph{exploration}, and summing them up is a natural way of dealing with the exploitation/exploration trade off.
The constant parameter $\alpha$ in $\xi_k(t)$ also controls this trade-off, and the theoretical analysis suggests to restrict to $\alpha\geq1/2$, and to choose $\alpha=1/2$ for optimal uniform performance (\ie, on all problems).

We present the regret bound, which was first stated in \cite{Auer02}, and to explain how Hoeffding's inequality is used we prove it below.

\begin{theorem}[Regret bound for $\UCB_1$]\label{thm:2:UCBregretBound}
    For simplicity, we restrict to $\alpha=2$.
    For any instance $I$ with $K$ arms with bounded distributions in $[0,1]$,
    of means $\mu_1,\dots,\mu_k$, and any horizon $T>10$,
    the $\UCB_1$ algorithm verifies the following finite-time regret bound
    \begin{equation}\label{eq:2:UCBregretBound}
        R_T^{\cA}(I) \leq \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left(1 + \frac{\pi^2}{3}\right)
        = \bigO{ \left(\sum_{k: \Delta_k>0} \frac{8}{\Delta_k}\right) \log(T)}.
    \end{equation}
\end{theorem}
%
\begin{proof}
    As we focus on bounded distributions in $[0,1]$, for which Hoeffding's inequality gives \eqref{eq:2:HoeffdingInequality}, as stated above in Lemma~\ref\label{lem:2:HoeffdingInequality}.
    Thanks to the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, we can focus on bounding $\E[N_k(T)]$ for any sub-optimal arm $k$.

    The first trick consists in writing the following, which is true for any $n \geq 1$:
    \begin{equation}\label{eq:2:trickInUCBProof}
        N_k(T) = 1 + \sum_{t=K+1}^T \indic(A(t) = k)
        \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n).
    \end{equation}

    Remember that $\UCB_a(t) = \widehat{\mu_a}(t) + \xi_a(t)$ for any arm $a$.
    According to the algorithm (see Algorithm~\ref{algo:2:indexPolicy} at line~6),
    arm $k$ is chosen if $\UCB_k(t) \geq \UCB_{k^*}(t)$.
    Both $\widehat{\mu_a}(t)$ and $\xi_a(t)$ use $N_k(t)$ samples of arm $a$,
    but as \eqref{eq:2:trickInUCBProof} above started to isolate $N_k(t)$, we want to add the freedom of considering different number of samples (using $n$), so we introduce the notation
    $\widehat{\mu_{a,n}}(t)$ and $\xi_{a,n}(t)$ where $N_k(t)$ is replaced with $n$ if $n \leq N_k(t)$,
    \ie, $\widehat{\mu_{a,n}}(t) = \frac{1}{n} \sum_{s=1}^{t-1} r_k(s) \indic(A(s)=k, N_k(s) \leq n)$ and $\xi_{a,n}(t) = \sqrt{2\log(t) / n}$.
    %
    Thus, we can continue from \eqref{eq:2:trickInUCBProof} and write
    \begin{align}\label{eq:2:decompositionInUCBProof}
        N_k(T) & \leq n + \sum_{t=n}^T \indic(A(t) = k, N_k(t) \geq n) \nonumber \\
        & \leq n + \sum_{t=n}^T \indic(\widehat{\mu_{k^*,N_{k^*}(t)}}(t) + \xi_{k^*,N_{k^*}(t)}(t) < \widehat{\mu_{k,N_k(t)}}(t) + \xi_{k,N_k(t)}(t), N_k(t) \geq n)  \nonumber \\
        & \leq n + \sum_{t=n}^T \indic(\min_{0 < n_{k^*} < t} \widehat{\mu_{k^*,n_{k^*}}}(t) + \xi_{k^*,n_{k^*}}(t) < \max_{n < n_k < t} \widehat{\mu_{k,n_k}}(t) + \xi_{k,n_k}(t)) \nonumber \\
        \intertext{So two union bounds on $n_{k^*}$ and $n_k$ give}
        & \leq n + \sum_{t=n}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n}^t  \indic( \widehat{\mu_{k^*,n_{k^*}}}(t) + \xi_{k^*,n_{k^*}}(t) < \widehat{\mu_{k,n_k}}(t) + \xi_{k,n_k}(t) )
    \end{align}

    Now, at time $t$, this event $\widehat{\mu_{k^*,n_{k^*}}}(t) + \xi_{k^*,n_{k^*}}(t) < \widehat{\mu_{k,n_k}}(t) + \xi_{k,n_k}(t)$
    implies at least one of the following three events:
    \begin{align*}
        I_1: & \widehat{\mu_{k^*,n_{k^*}}}(t) \leq \mu_{k^*} - \xi_{k^*,n_{k^*}}(t),
        I_2: & \widehat{\mu_{k,n_k}}(t) \geq \mu_k + \xi_{k,n_k}(t),
        I_3: & \mu_{k^*} < \mu_k + 2 \xi_{k,n_k}(t).
    \end{align*}

    For the first two events $I_1$ and $I_2$, we carefully fixed the number of samples so they are no longer random (they are respectively to $n_{k^*}$ and $n_k$), we can use Hoeffding's inequality to obtain $\Pr(I_j) \leq \frac{1}{t^{4}}$ for $j=1,2$.

    For the last event $I_3$,
    observe that $\mu_{k^*} < \mu_k + 2 \xi_{k,n_k}(t)$ means $\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0$, where we remind the notation of the gap $\Delta_k = \mu_{k^*} - \mu_k$.
    Thus if $n_k \geq \lceil \frac{8 \log(T)}{\Delta_k^2} \rceil \geq \lceil \frac{8 \log(t)}{\Delta_k^2} \rceil$,
    then this inequality actually cannot happen:
    $\Pr(\Delta_k - 2 \sqrt{2\log(t) / n_k} < 0) = 0$,
    so as soon as $n_k \geq \frac{8 \log(T)}{\Delta_k^2}$, $\Pr(I_3) = 0$.

    Thus we can take the expectation of $N_k(T)$ from equation~\eqref{eq:2:decompositionInUCBProof}, and continue by setting
    with $n = n_T = \lceil \frac{8 \log(T)}{\Delta_k^2} \rceil$:

    \begin{align}\label{eq:2:decompositionInUCBProof2}
        \E[ N_k(T) ]
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  \Pr( \widehat{\mu_{k^*,n_{k^*}}}(t) + \xi_{k^*,n_{k^*}}(t) < \widehat{\mu_{k,n_k}}(t) + \xi_{k,n_k}(t) ) \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  \Pr( I_1 ) + \Pr( I_2 ) + \Pr( I_3 ) \nonumber \\
        & \leq 1 + \frac{8 \log(T)}{\Delta_k^2} + \sum_{t=n_T}^T \sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  2 \frac{1}{t^4} \nonumber \\
        \intertext{We actually have that $\sum_{n_{k^*}=1}^t \sum_{n_{k}=n_T}^t  2 \frac{1}{t^4} \leq 2 \sum_{n_{k^*}=1}^t \sum_{n_{k}=1}^t \frac{1}{t^4} \leq 2 \sum_{n_{k^*}=1}^t \frac{1}{t^3} \leq 2 \frac{1}{t^2}$}
        & \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^T \frac{1}{t^2} \nonumber \\
        & \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \sum_{t=1}^{\infty} \frac{1}{t^2} \nonumber \\
        \E[ N_k(T) ]
        & \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + 2 \frac{\pi^2}{6}.
    \end{align}

    To sum up, we obtain the following finite-time bound on $N_k(T)$ for any sub-optimal arm $k$:
    \begin{align*}
        \E[N_k(T)] \leq \frac{8 \log(T)}{\Delta_k^2} + 1 + \frac{\pi^2}{3}.
    \end{align*}
    %
    By using the regret decomposition from Lemma~\ref{lem:2:RegretDecomposition}, this gives
    \begin{align*}
        R_T^{\UCB_1} &\leq \sum_{k: \Delta_k>0} \Delta_k \E[N_k(T)] \\
        & \leq \frac{8}{\Delta_k} \log(T) + \sum_{k: \Delta_k>0} \Delta_k \left( 1 + \frac{\pi^2}{3} \right)
        = \frac{8}{\Delta_k} \log(T) + \smallO{\log(T)}.
    \end{align*}

    More details on this proof can be found for instance in Section~1.3 of \cite{Slivkins2019}, while for instance Chapter~2 of \cite{Bubeck12} Theorem~2.1) and Chapter~7 of \cite{LattimoreBanditAlgorithmsBook} (Theorem~7.2) both give more general proofs (\eg, for any $\alpha\geq\frac{1}{2}$).
\end{proof}

What this theorem shows is that the $\UCB_1$ algorithm achieves order-optimal problem-dependent bounds with a $\cO(K\log(T)/\Delta^2)$ regret (if $\Delta=\min\limits_{k: \Delta_k>0} \Delta$).
The $\UCB_1$ algorithm has the advantage of not relying on any prior knowledge of the problem difficulty, of being anytime, and computationally simple, for both memory and time aspects.
It requires a constant storage capacity for each arm (storing $X_k(t)$ and $N_k(t)$ require two integers or a \texttt{float} and an integer, \ie, $\bigO{1}$), giving a storage capacity of $\bigO{K}$, independently of the horizon $T$.
It also needs to perform some mathematical operations, but a square-root and a logarithm are cheap, thus it costs a $\bigO{1}$ constant time for each arm and time step, thus a time $\bigO{K T}$ for the $T$ time steps.
%
We use it as the base building block for other policies in Chapter~\ref{chapter:4}, for these reasons.
Moreover, the $\UCB_1$ algorithm has the advantage of being easy to explain and understand, and its decisions can be interpreted clearly: by reading the values of $\widehat{\mu_k}(t)$ and $\UCB_k(t)$, anyone can see why an arm was chosen at anytime.
This is why it is used in our demonstration presented in Section~\ref{sec:4:gnuradio}.
